### 2019-01-15 submission/manuscript.tex
Fresh make write.paper

\hypertarget{assessing-the-differences-in-16s-rrna-gene-sequencing-due-to-high-fidelity-dna-polymerase}{%
\section{Assessing the Differences in 16S rRNA Gene Sequencing Due to
High Fidelity DNA
Polymerase}\label{assessing-the-differences-in-16s-rrna-gene-sequencing-due-to-high-fidelity-dna-polymerase}}

\hypertarget{influence-of-dna-polymerase-on-pcr-bias-chimera-formation-and-errors-when-sequencing-16s-rrna-genes-from-complex-communities}{%
\section{Influence of DNA Polymerase on PCR bias, chimera formation, and
errors when sequencing 16S rRNA genes from complex
communities}\label{influence-of-dna-polymerase-on-pcr-bias-chimera-formation-and-errors-when-sequencing-16s-rrna-genes-from-complex-communities}}

\begin{center}
\vspace{25mm}

@@

Co-author e-mails:

\begin{itemize}
\tightlist
\item
  \href{mailto:marcsze@med.umich.edu}{\nolinkurl{marcsze@med.umich.edu}}
\end{itemize}

@@

@@

\textbf{\emph{Sequencing errors vary by the number of cycles and the DNA
polymerase used in PCR.}} The presence of sequence errors can confound
the ability to accurately classify 16S rRNA gene sequences and group
sequences into operational taxonomic units (OTUs). More importantly,
sequencing errors themselves can alter the representation of the
community. Therefore, it is important to minimize the number of
sequencing errors. Using a widely-used approach that generates the
lowest reported error rate, we quantified the error rate by sequencing
the V4 region of the 16S rRNA genes from an 8 member mock community. We
also removed any contigs that were at least three bases more similar to
a chimera of two references than to a single reference sequence
{[}ref{]}. Regardless of the polymerase, the error rate increased with
the number of rounds of amplification. Using 30 rounds of PCR is a
common approach across diverse types of samples. Among the data
generated using 30 rounds of PCR the Accuprime polymerase had the
highest error rate (i.e.~0.124\%) followed by the Platinum
(i.e.~0.094\%), Phusion (i.e.~0.064\%), Kappa (i.e.~0.062\%), and Q5
(i.e.~0.060\%) polymerases (Figure 1). When we applied the
pre.clustering denoising step, which merges the counts of reads within 2
nt of a more abundant sequence, the error rates dropped considerably
such that the Platinum polymerase had the highest error rate
(i.e.~0.014\%) followed by the Accuprime (i.e.~0.012\%), Q5
(i.e.~0.0053\%), Phusion (i.e.~0.0049\%), and Kappa (i.e.~0.0049\%)
polymerases (Figure 1). Although specific recommendations are difficult
to make because the impact of the actual community structure and
concentration of the initial DNA template are likely to have an impact
on the results, it is clear that using as few PCR cycles as necessary
and a polymerase with the lowest possible error rate is a good guide to
minimizing the impact of polymerase on the error rate.

\textbf{\emph{The fraction of sequences identified as being chimeric
varies by the number of cycles and the DNA polymerase used in PCR.}}
Chimeric PCR products can significantly confound downstream analyses.
Although numerous bioinformatic tools exist to identify and remove
chimeric sequences with high specificity, their sensitivity is
relatively low and can be reduced by the presence of sequencing errors.
We identified those contigs that were at least three bases more similar
to a chimera of two references than to a single reference sequence. As
expected from previous studies, the number of chimeras increased with
rounds of amplification. Interestingly, the fraction of chimeric
sequences from the mock community varied by the type of polymerase used.
After 30 rounds of PCR, the Platinum polymerase had the highest chimera
rate (i.e.~18.2\%) followed by the Q5 (i.e.~8.1\%), Phusion
(i.e.~7.5\%), Kappa (i.e.~2.3\%), and Accuprime (i.e.~0.9\%)
polymerases. We used the UChime algorithm to detect chimeras in our
pre-clustered mock community data and calculate the algorithm's
sensitivity and specificity. For all polymerases, the specificity was
above 95.8\% and showed a weak association with the number of cycles
used. There was considerable inter-polymerase and inter-round of
amplification variation in the sensitivity of UChime to detect the
chimeras from the mock community. This suggested that the residual error
rate after pre-clustering the sequence data did not compromise the
sensitivity. The sensitivity of UChime varied between 50 and 84.8\%. The
generalizability of these results is limited because we used a single
mock community with limited genetic diversity. Although we did not know
the true chimera rate for our four human stool samples, we were able to
calculate the fraction of sequences that UChime identified as being
chimeric. These results followed those from the mock communities;
additional rounds of amplification significantly increased the rate of
chimeras and there was variation between the polymerases that we used.
Although it was not possible to identify the features of a polymerase
that resulted in higher rates of chimeras, it is clear that using the
smallest number of PCR cycles possible will minimize the impacts of
chimeras.

\textbf{\emph{At the sequence level, PCR amplification bias is subtle.}}
Since researchers began using PCR to sequence 16S rRNA genes there has
been concern that amplifying fragments from a mixed template pool could
lead to a biased representation in the pool of products and would
confound downstream analyses. The mock community was generated by mixing
equal amounts of genomic DNA from each of the 8 bacteria resulting in
uneven representation of the \emph{rrn} operons across the bacteria. The
vendor of the mock community subjects each lot of genomic DNA to shotgun
sequencing to more accurately quantify the actual abundance of each
organism in the community. We compared the expected relative abundance
of the 16S rRNA genes from each bacterium in the mock community to the
data we generated across rounds of amplification and polymerase.
Interestingly, for some bacteria, their representation became less
biased with additional rounds of PCR (e.g.~\emph{L. fermentum}), while
others became more biased (e.g.~\emph{E. faecalis}), and others had
little change (e.g.~\emph{B. subtilis}). The percentage of bases in the
V4 region that were guanines or cytosines was not predictive of the
amount of bias. Across the strains there was no variation in the length
of their V4 regions and they each had the same sequence in the region
that the primers annealed. One of the bacteria represented in the mock
community, \emph{S. enterica}, had 6 identical copies of the V4 region
and 1 copy that differed from those by one nucleotide. The dominant copy
had a thymidine and the rare copy had a guanine. We used the sequence
data to calculate the ratio of the dominant to rare variants from
\emph{S. enterica} (Figure XXX). The Accuprime, Phusion, Platinum, and
Q5 polymerases converged to a ratio of 5.4; however, the ratio for the
Kappa polymerase varied between 6.1 and 6.8 for 25 to 35 rounds of PCR.
Given the subtle nature of the variation in the relative abundances of
each 16S rRNA gene fragment, it was not possible to create generalizable
rules that would explain the bias and the previous observation regarding
the guanine and cytosine relative abundance was not observed with our
data.

\textbf{\emph{At the community level, the effects of PCR amplification
bias grow with additional rounds of PCR.}} Because the distance between
samples and across rounds of PCR could be artificially inflated due to
sequencing errors and chimeras, we used analyzed the alpha and beta
diversity of the mock community data at different phases of the sequence
curation pipeline. First, we removed the chimeras from the mock
community data as described above and mapped the individual reads to the
OTUs that the 16S rRNA gene fragments would cluster into if there were
no sequencing errors. This gave us a community distribution that
reflected the distribution following PCR without any artifacts. Alpha
diversity metrics such as richness and Shannon and Inverse Simpson
diversity metrics increased with the number of rounds of PCR for all
polymerases except for the Kappa polymerase, where the richness remained
constant but the diversity decreased. These data suggest that PCR had
the effect of making the community distribution more even than it was
originally. Next, we used the true sequence diversity, but removed
chimeras based on mapping reads to all possible chimeras between
reference sequences, and clustered the reads to OTUs. The same trends
were observed as with perfect sequence data except that the richness and
diversity metrics trended higher. Finally, we used the observed sequence
data and used the UCHIME algorithm to identify chimeras. Again, we
observed the same trends as with the other analyses except the richness
and diversity metrics trended higher. These comparisons demonstrated
that although the bias at the sequence level was subtle, PCR does have a
bias that is exacerbated by errors and chimeras. When we measured the
Bray-Curtis distance between the communities observed after 25 rounds of
amplification and those at 30 and 35 rounds of amplification the
distances between 25 and 35 rounds were higher than between 25 and 30
rounds for each of the polymerases by an average of 0.023 units. The
Platinum polymerase varied the most across rounds of amplification (25
vs 30 rounds: 0.13; 25 vs 35 rounds: 0.16). Although the distances
between samples were small, the ordination of these distances showed a
clear change in community structure with increasing rounds of PCR. This
observation was supported by our statistical analysis, which revealed
that the number of rounds of PCR (R\textsuperscript{2}=0.20,
P\textless{}0.001) was a larger factor than the choice of polymerase
(R\textsuperscript{2}=0.20, P\textless{}0.001). These results
demonstrate that subtle differences in relative abundances can have an
impact on overall community structure. This variation underscores the
importance of only comparing sequence data that have been generated
using the same PCR conditions.

\textbf{\emph{The choice of polymerase or the number of rounds of
amplification have little impact on the relative interpretation of
community-wide metrics of diversity.}} The biases that we observed at
the population and community levels using mock community data appeared
to be small relative to the expected differences between biological
samples. To study this further, we calculated alpha and beta-diversity
metrics using the human stool samples for each of the polymerases and
rounds of amplification. We calculated the number of observed OTUs,
Shannon diversity, inverse Simpson diversity index for each condition
and donor. Although there were clear differences between conditions, the
relative ordering of the stool samples did not meaningfully vary across
conditions. When we characterized the variation between rounds of
amplification using human stool samples, the distance between the 25 and
30 rounds and 25 and 35 rounds varied considerably between samples and
polymerases. In general the inter-round variation was lowest for the
data generated using the Kappa and Accuprime polymerases. The data
generated using the Platinum polymerase was consistent across rounds,
but it was more biased than these polymerases. Considering the average
distance across the four samples varied between 0.39 and 0.56,
regardless of the polymerases and number of rounds of amplification, any
bias due to amplification is unlikely to obscure community-wide
differences between samples. In support of this was our principle
coordinates analysis of the Bray-Curtis distances, which revealed
distinct clusters by donor. Within each cluster there were no obvious
patterns related to the polymerase or number of rounds of PCR. However,
our statistical analysis using adonis revealed statistically significant
differences in the community structures with the subject explaining the
most variation (R\textsuperscript{2}=0.79, P\textless{}0.001), followed
by the number of rounds of PCR (R\textsuperscript{2}=0.044,
P\textless{}0.001) and the choice of polymerase
(R\textsuperscript{2}=0.033, P\textless{}0.001). Together, these results
indicate that for a coarse analysis of communities, the choice of number
of rounds of amplification or polymerase are not important, but that
they must be consistent across samples. It is difficult to develop a
specific recommendation based on the level of bias across rounds of PCR
or polymerases; however, the general suggestion is to use as few rounds
of amplification as possible.

\textbf{\emph{There is little evidence of a relationship between
polymerase or number of rounds of amplification on PCR drift.}} There
have been concerns that the same template DNA subjected to the same PCR
conditions could result in different representations of communities
because of random drift. To test this, we determined the average
Bray-Curtis distance between replicate reactions using the same
polymerase and number of rounds of amplification. Using the mock
community data there were no obvious trends. The average Bray-Curtis
distance within a set of conditions varied by 0.063 to 0.11 units.
Although we did not obtain replicates of each of the stool samples, the
intra-sample variation for each set of conditions was consistent and
varied between 0.50 and 0.56 units. These data suggest that amplicon
sequencing is robust to random variation in amplification and that any
differences are likely to be smaller than what is considered
biologically relevant.

@@

The observed sequencing error rates and alpha diversity metrics followed
the manufacturers' measurements of their polymerases' fidelity.
Accuprime and Platinum have fidelity that are approximately 10-times
higher than that of Taq whereas the fidelity of Phusion, Q5, and Kappa
are more than 100 times higher. Among these polymerases, the Kappa
polymerase resulted in the the lowest error rate, lowest chimera rate,
and least bias across rounds of PCR. Considering polymerase development
is an active area of commercial development and it likely that new
polymerases will come to market, it is important for researchers to
understand how changing the polymerase impacts downstream analyses for
their type of samples.

Over the past 20 years, a large literature has attempted to document
various PCR biases and underscores the fact that data based on
amplification of DNA from a mixed community are not a true
representation of the actual community. In addition to obvious biases
imposed by primer selection, other factors inherent in PCR can influence
the representation of communities. Factors that can lead to preferential
amplification of one fragment over another have included guanine and
cytosine composition, length, flanking DNA composition, amount of DNA
shearing, and number of rounds of PCR (24, 27--33). In addition,
environmental and reagent contaminants can also have a significant
impact on the analysis of low biomass samples (38). Less well understood
is the effect of phylogenetic diversity on bias and chimera formation.
Communities with low phylogenetic diversity may be more prone to chimera
formation since chimeras are more likely to form among closely related
sequences (14, 35). The interaction of these various influences on PCR
artifacts are complex and difficult to tease apart. Minimizing the level
of DNA shearing and using the fewest number of rounds of PCR with a
polymerase that has the highest possible fidelity are strategies that
can be employed to minimize the formation of chimeras. Although care
should always be taken when choosing a polymerase for 16S rRNA gene
sequencing, our observations show that the differences between a variety
of polymerases are smaller than the actual biological variation in fecal
communities between individuals. Therefore, if the biological signal of
interest is similar to differences in fecal bacterial communities found
between individuals, then the type of high fidelity DNA polymerase used
will only minimally change the results.

Even with these strategies it is impossible to remove all PCR artifacts.
Beyond the imperfections of the best polymerases, sometimes difficult to
lyse organisms require stringent lysis steps and low biomass samples
require additional rounds of PCR. A host of bioinformatics tools are
available for removing residual sequencing errors (18, 39--41). These
tools struggle to correctly differentiate true biological diversity
(i.e.~Amplicon Sequencing Variants or ASVs) and PCR or sequencing
errors. Although these methods remove residual errors, they also risk
splitting intragenomic variants into separate ASVs, merging 16S rRNA
gene sequences from different taxa into the same ASV. Other tools are
available for removing chimeras (14, 35) where there is a trade off
between the sensitivity of detecting chimeras and the specificity of
correctly calling a sequence a chimera. In recent years, parameters for
these algorithms have been changed to increase their sensitivity with
little evaluation of the effects on the specificity of the algorithms
(39, 41). Others recommend removing any read that has an abundance below
a specified threshold (e.g.~removing all sequences that only appear
once) (20, 39--41). This method must be approached with caution as such
approaches are likely to introduce a different bias of the community
representation and ignore the fact that artifacts may be quite abundant.
Ultimately, researchers must test their hypotheses with multiple methods
to validate the claims they reach with any one method (42). All methods
have biases and limitations and we must use complementary methods to
obtain robust results.

@@

\textbf{\emph{Human samples.}} Fecal samples were obtained from 4
individuals who were part of an earlier study (43). These samples were
collected using a protocol approved by the University of Michigan
Institutional Review Board. For this study, the samples were
de-identified. DNA was extracted from the fecal samples using the
MOBIO\textsuperscript{TM} PowerMag Microbiome RNA/DNA extraction kit
(now Qiagen, MD, USA).

@@

\textbf{\emph{Sequence processing.}} The mothur software program (v
1.41) was used for all sequence processing steps (44). The protocol has
been previously published (18)
(\url{https://www.mothur.org/wiki/MiSeq_SOP}). Briefly, paired reads
were assembled using mothur's make.contigs command to correct errors
introduced by sequencing (18). Any assembled contigs that contained an
ambiguous base call, mapped to the incorrect region of the 16S rRNA
gene, or appeared to be a contaminant were removed from subsequent
analyses. Sequences were further denoised using mothur's pre.cluster
command to merge the counts of sequences that were within 2 nt of a more
abundant sequence. The VSEARCH implementation of UCHIME was used to
screen for chimeras (35, 45). At various stages in the sequence
processing pipeline for the mock community data, the mothur seq.error
command was used to quantify the sequencing error rate as well as the
true chimera rate. This command uses the true sequences from the mock
community to generate all possible chimeras and removes any contigs that
were at least three bases more similar to a chimera than to a reference
sequence. The command then counts the number of substitutions,
insertions, and deletions in the contig relative to the reference
sequence and reports the error rate without the inclusion of chimeric
sequences (19). The reference sequences and operon copy number for each
bacterium were obtained from the ZymoBIOMICS\textsuperscript{TM}
Microbial Community DNA Standard protocol
(\url{https://web.archive.org/web/20181221151905/https://www.zymoresearch.com/media/amasty/amfile/attach/_D6305_D6306_ZymoBIOMICS_Microbial_Community_DNA_Standard_v1.1.3.pdf}).
Sequences were assigned to operational taxonomic units (OTUs) at a
threshold of 3\% dissimilarity using the OptiClust algorithm (46). To
adjust for unequal sequencing when measuring alpha and beta diversity,
all samples were rarefied to 1,000 sequences 1,000 times for downstream
analysis. The Good's coverage for the samples at this level of sampling
was routinely greater than 95\%.

\textbf{\emph{Statistical analysis.}} All analysis was done with the R
(v 3.5.0) software package (47). Data transformation and graphing were
completed using the tidyverse package (v 1.2.1) (48). The distance
matrix data was analyzed using the adonis function within the vegan
package (v 2.5.3) (49).

@@

\hypertarget{figure-lengends}{%
\subsection{Figure Lengends}\label{figure-lengends}}

>>>>

\vspace{25mm}
\begin{center}

\textbf{\huge Assessing the differences in 16S rRNA gene sequence data that are due to choice of DNA polymerase and number of rounds of PCR}

\vspace{50mm}

\textbf{Running title:} Quantifying the choice of DNA polymerase

\vspace{25mm}

@@

@@

PCR amplification of 16S rRNA genes is a critical, yet under appreciated
step in the generation of sequence data to describe the taxonomic
composition of microbial communities. Numerous factors in the design of
PCR can impact the sequencing error rate, the abundance of chimeric
sequences, and the degree to which the fragments in the product
represent their abundance in the original sample. The choice of PCR
cycling conditions are known to impact the outcome of PCR and downstream
inferences. We compared the performance of high fidelity polymerases and
varying number of rounds of amplification when amplifying a mock
community with known sequences and community structure and human stool
samples. Although it is impossible to derive specific recommendations,
we did observe general trends. Namely, using a polymerase with the
highest possible fidelity and minimizing the number of rounds of PCR
reduced the sequencing error rate, fraction of chimeric sequences, and
bias. Evidence of bias at the sequence level was subtle and could not be
ascribed to the fragments' fraction of bases that were guanines or
cytosines. When analyzing the mock community data, the amount that the
community deviated from the expected composition increased with rounds
of PCR. The effect was inconsistent for human stool samples. Overall the
results underscore the difficulty of comparing sequence data that are
generated by different PCR protocols; however, the results indicate that
the variation in human stool samples is generally larger than that
introduced by the choice of polymerase or number of rounds of PCR.

\hypertarget{importance}{%
\subsubsection{Importance}\label{importance}}

The growth in interest in characterizing microbial communities from
diverse communities has exploded with the steep decline in sequencing
costs. A significant amount of effort has gone into understanding the
error profiles of DNA sequencers; however, little work has been done to
understand the effects of choices that researchers make to generate the
PCR amplicons that are sequenced. Here we quantified the effects of the
choice of polymerase and number of PCR cycles on the quality of
downstream data. Overall, we found that these choices can have a
profound impact on the way that a microbial community is represented in
the sequence data. Although the effects are relatively small compared to
the variation in human stool samples, care should be taken to use
polymerases with the highest possible fidelity and to minimize the
number of rounds of PCR.

@@

\textbf{\emph{Sequencing errors vary by the number of cycles and the DNA
polymerase used in PCR.}} The presence of sequence errors can confound
the ability to accurately classify 16S rRNA gene sequences and group
sequences into operational taxonomic units (OTUs). More importantly,
sequencing errors themselves can alter the representation of the
community. Therefore, it is important to minimize the number of
sequencing errors. Using a widely-used approach that generates the
lowest reported error rate, we quantified the error rate by sequencing
the V4 region of the 16S rRNA genes from an 8 member mock community. We
also removed any contigs that were at least three bases more similar to
a chimera of two references than to a single reference sequence (18, 19,
38). Regardless of the polymerase, the error rate increased with the
number of rounds of amplification (Figure 1). Using 30 rounds of PCR is
a common approach across diverse types of samples. Among the data
generated using 30 rounds of PCR the Accuprime polymerase had the
highest error rate (i.e.~0.124\%) followed by the Platinum
(i.e.~0.094\%), Phusion (i.e.~0.064\%), Kappa (i.e.~0.062\%), and Q5
(i.e.~0.060\%) polymerases (Figure 1). When we applied the
pre.clustering denoising step, which merges the counts of reads within 2
nt of a more abundant sequence, the error rates dropped considerably
such that the Platinum polymerase had the highest error rate
(i.e.~0.014\%) followed by the Accuprime (i.e.~0.012\%), Q5
(i.e.~0.0053\%), Phusion (i.e.~0.0049\%), and Kappa (i.e.~0.0049\%)
polymerases (Figure 1). Although specific recommendations are difficult
to make because the impact of the actual community structure and
concentration of the initial DNA template are likely to have an impact
on the results, it is clear that using as few PCR cycles as necessary
and a polymerase with the lowest possible error rate is a good guide to
minimizing the impact of polymerase on the error rate.

\textbf{\emph{The fraction of sequences identified as being chimeric
varies by the number of cycles and the DNA polymerase used in PCR.}}
Chimeric PCR products can significantly confound downstream analyses.
Although numerous bioinformatic tools exist to identify and remove
chimeric sequences with high specificity, their sensitivity is
relatively low and can be reduced by the presence of sequencing errors.
We identified those contigs that were at least three bases more similar
to a chimera of two references than to a single reference sequence. As
expected from previous studies, the number of chimeras increased with
rounds of amplification (Figure 2A). Interestingly, the fraction of
chimeric sequences from the mock community varied by the type of
polymerase used. After 30 rounds of PCR, the Platinum polymerase had the
highest chimera rate (i.e.~18.2\%) followed by the Q5 (i.e.~8.1\%),
Phusion (i.e.~7.5\%), Kappa (i.e.~2.3\%), and Accuprime (i.e.~0.9\%)
polymerases. We used the UChime algorithm to detect chimeras in our
pre-clustered mock community data and calculate the algorithm's
sensitivity and specificity (Figure 2A). For all polymerases, the
specificity was above 95.7\% and showed a weak association with the
number of cycles used. There was considerable inter-polymerase and
inter-round of amplification variation in the sensitivity of UChime to
detect the chimeras from the mock community. This suggested that the
residual error rate after pre-clustering the sequence data did not
compromise the sensitivity. The sensitivity of UChime varied between 50
and 87.0\%. The generalizability of these results is limited because we
used a single mock community with limited genetic diversity. Although we
did not know the true chimera rate for our four human stool samples, we
were able to calculate the fraction of sequences that UChime identified
as being chimeric (Figure 2B). These results followed those from the
mock communities; additional rounds of amplification significantly
increased the rate of chimeras and there was variation between the
polymerases that we used. Although it was not possible to identify the
features of a polymerase that resulted in higher rates of chimeras, it
is clear that using the smallest number of PCR cycles possible will
minimize the impacts of chimeras.

\textbf{\emph{At the sequence level, PCR amplification bias is subtle.}}
Since researchers began using PCR to sequence 16S rRNA genes there has
been concern that amplifying fragments from a mixed template pool could
lead to a biased representation in the pool of products and would
confound downstream analyses. The mock community was generated by mixing
equal amounts of genomic DNA from each of the 8 bacteria resulting in
uneven representation of the \emph{rrn} operons across the bacteria. The
vendor of the mock community subjects each lot of genomic DNA to shotgun
sequencing to more accurately quantify the actual abundance of each
organism in the community. We compared the expected relative abundance
of the 16S rRNA genes from each bacterium in the mock community to the
data we generated across rounds of amplification and polymerase (Figure
3). Interestingly, for some bacteria, their representation became less
biased with additional rounds of PCR (e.g.~\emph{L. fermentum}), while
others became more biased (e.g.~\emph{E. faecalis}), and others had
little change (e.g.~\emph{B. subtilis}). The percentage of bases in the
V4 region that were guanines or cytosines was not predictive of the
amount of bias. Across the strains there was no variation in the length
of their V4 regions and they each had the same sequence in the region
that the primers annealed. One of the bacteria represented in the mock
community, \emph{S. enterica}, had 6 identical copies of the V4 region
and 1 copy that differed from those by one nucleotide. The dominant copy
had a thymidine and the rare copy had a guanine. We used the sequence
data to calculate the ratio of the dominant to rare variants from
\emph{S. enterica} (Figure S1). The Accuprime, Phusion, Platinum, and Q5
polymerases converged to a ratio of 5.4; however, the ratio for the
Kappa polymerase varied between 6.1 and 6.8 for 25 to 35 rounds of PCR.
Given the subtle nature of the variation in the relative abundances of
each 16S rRNA gene fragment, it was not possible to create generalizable
rules that would explain the bias and the previous observation regarding
the guanine and cytosine relative abundance was not observed with our
data.

\textbf{\emph{At the community level, the effects of PCR amplification
bias grow with additional rounds of PCR.}} Because the distance between
samples and across rounds of PCR could be artificially inflated due to
sequencing errors and chimeras, we used analyzed the alpha and beta
diversity of the mock community data at different phases of the sequence
curation pipeline (Figure 4). First, we removed the chimeras from the
mock community data as described above and mapped the individual reads
to the OTUs that the 16S rRNA gene fragments would cluster into if there
were no sequencing errors. This gave us a community distribution that
reflected the distribution following PCR without any artifacts. Alpha
diversity metrics such as richness and Shannon and Inverse Simpson
diversity metrics increased with the number of rounds of PCR for all
polymerases except for the Kappa polymerase, where the richness remained
constant but the diversity decreased (Figure 4A). These data suggest
that PCR had the effect of making the community distribution more even
than it was originally. Next, we used the true sequence diversity, but
removed chimeras based on mapping reads to all possible chimeras between
reference sequences, and clustered the reads to OTUs. The same trends
were observed as with perfect sequence data except that the richness and
diversity metrics trended higher (Figure 4A). Finally, we used the
observed sequence data and used the UCHIME algorithm to identify
chimeras. Again, we observed the same trends as with the other analyses
except the richness and diversity metrics trended higher (Figure 4A).
These comparisons demonstrated that although the bias at the sequence
level was subtle, PCR does have a bias that is exacerbated by errors and
chimeras. When we measured the Bray-Curtis distance between the
communities observed after 25 rounds of amplification and those at 30
and 35 rounds of amplification the distances between 25 and 35 rounds
were higher than between 25 and 30 rounds for each of the polymerases by
an average of 0.022 units (Figure 4B). The Platinum polymerase varied
the most across rounds of amplification (25 vs 30 rounds: 0.12; 25 vs 35
rounds: 0.15). Although the distances between samples were small, the
ordination of these distances showed a clear change in community
structure with increasing rounds of PCR (Figure 4C). This observation
was supported by our statistical analysis, which revealed that the
number of rounds of PCR (R\textsuperscript{2}=0.21, P\textless{}0.001)
was a larger factor than the choice of polymerase
(R\textsuperscript{2}=0.20, P\textless{}0.001). These results
demonstrate that subtle differences in relative abundances can have an
impact on overall community structure. This variation underscores the
importance of only comparing sequence data that have been generated
using the same PCR conditions.

\textbf{\emph{The choice of polymerase or the number of rounds of
amplification have little impact on the relative interpretation of
community-wide metrics of diversity.}} The biases that we observed at
the population and community levels using mock community data appeared
to be small relative to the expected differences between biological
samples. To study this further, we calculated alpha and beta-diversity
metrics using the human stool samples for each of the polymerases and
rounds of amplification. We calculated the number of observed OTUs,
Shannon diversity, inverse Simpson diversity index for each condition
and donor (Figure 5A). Although there were clear differences between
conditions, the relative ordering of the stool samples did not
meaningfully vary across conditions. When we characterized the variation
between rounds of amplification using human stool samples, the distance
between the 25 and 30 rounds and 25 and 35 rounds varied considerably
between samples and polymerases (Figure 5B). In general the inter-round
variation was lowest for the data generated using the Kappa and
Accuprime polymerases. The data generated using the Platinum polymerase
was consistent across rounds, but it was more biased than these
polymerases. Considering the average distance across the four samples
varied between 0.39 and 0.56, regardless of the polymerases and number
of rounds of amplification, any bias due to amplification is unlikely to
obscure community-wide differences between samples. In support of this
was our principle coordinates analysis of the Bray-Curtis distances,
which revealed distinct clusters by donor (Figure 5C). Within each
cluster there were no obvious patterns related to the polymerase or
number of rounds of PCR. However, our statistical analysis using adonis
revealed statistically significant differences in the community
structures with the subject explaining the most variation
(R\textsuperscript{2}=0.79, P\textless{}0.001), followed by the number
of rounds of PCR (R\textsuperscript{2}=0.044, P\textless{}0.001) and the
choice of polymerase (R\textsuperscript{2}=0.033, P\textless{}0.001).
Together, these results indicate that for a coarse analysis of
communities, the choice of number of rounds of amplification or
polymerase are not important, but that they must be consistent across
samples. It is difficult to develop a specific recommendation based on
the level of bias across rounds of PCR or polymerases; however, the
general suggestion is to use as few rounds of amplification as possible.

\textbf{\emph{There is little evidence of a relationship between
polymerase or number of rounds of amplification on PCR drift.}} There
have been concerns that the same template DNA subjected to the same PCR
conditions could result in different representations of communities
because of random drift. To test this, we determined the average
Bray-Curtis distance between replicate reactions using the same
polymerase and number of rounds of amplification (Figure 6). Using the
mock community data there were no obvious trends. The average
Bray-Curtis distance within a set of conditions varied by 0.063 to 0.11
units. Although we did not obtain replicates of each of the stool
samples, the intra-sample variation for each set of conditions was
consistent and varied between 0.50 and 0.56 units. These data suggest
that amplicon sequencing is robust to random variation in amplification
and that any differences are likely to be smaller than what is
considered biologically relevant.

@@

The observed sequencing error rates and alpha diversity metrics followed
the manufacturers' measurements of their polymerases' fidelity (Figure
1). Accuprime and Platinum have fidelity that are approximately 10-times
higher than that of Taq whereas the fidelity of Phusion, Q5, and Kappa
are more than 100 times higher. Among these polymerases, the Kappa
polymerase resulted in the the lowest error rate, lowest chimera rate,
and least bias across rounds of PCR. Considering polymerase development
is an active area of commercial development and it likely that new
polymerases will come to market, it is important for researchers to
understand how changing the polymerase impacts downstream analyses for
their type of samples.

Over the past 20 years, a large literature has attempted to document
various PCR biases and underscores the fact that data based on
amplification of DNA from a mixed community are not a true
representation of the actual community. In addition to obvious biases
imposed by primer selection, other factors inherent in PCR can influence
the representation of communities. Factors that can lead to preferential
amplification of one fragment over another have included guanine and
cytosine composition, length, flanking DNA composition, amount of DNA
shearing, and number of rounds of PCR (24, 27--33). In addition,
environmental and reagent contaminants can also have a significant
impact on the analysis of low biomass samples (39). Less well understood
is the effect of phylogenetic diversity on bias and chimera formation.
Communities with low phylogenetic diversity may be more prone to chimera
formation since chimeras are more likely to form among closely related
sequences (14, 35). The interaction of these various influences on PCR
artifacts are complex and difficult to tease apart. Minimizing the level
of DNA shearing and using the fewest number of rounds of PCR with a
polymerase that has the highest possible fidelity are strategies that
can be employed to minimize the formation of chimeras. Although care
should always be taken when choosing a polymerase for 16S rRNA gene
sequencing, our observations show that the differences between a variety
of polymerases are smaller than the actual biological variation in fecal
communities between individuals. Therefore, if the biological signal of
interest is similar to differences in fecal bacterial communities found
between individuals, then the type of high fidelity DNA polymerase used
will only minimally change the results.

Even with these strategies it is impossible to remove all PCR artifacts.
Beyond the imperfections of the best polymerases, sometimes difficult to
lyse organisms require stringent lysis steps and low biomass samples
require additional rounds of PCR. A host of bioinformatics tools are
available for removing residual sequencing errors (18, 40--42). These
tools struggle to correctly differentiate true biological diversity
(i.e.~Amplicon Sequencing Variants or ASVs) and PCR or sequencing
errors. Although these methods remove residual errors, they also risk
splitting intragenomic variants into separate ASVs, merging 16S rRNA
gene sequences from different taxa into the same ASV. Other tools are
available for removing chimeras (14, 35) where there is a trade off
between the sensitivity of detecting chimeras and the specificity of
correctly calling a sequence a chimera. In recent years, parameters for
these algorithms have been changed to increase their sensitivity with
little evaluation of the effects on the specificity of the algorithms
(40, 42). Others recommend removing any read that has an abundance below
a specified threshold (e.g.~removing all sequences that only appear
once) (20, 40--42). This method must be approached with caution as such
approaches are likely to introduce a different bias of the community
representation and ignore the fact that artifacts may be quite abundant.
Ultimately, researchers must test their hypotheses with multiple methods
to validate the claims they reach with any one method (43). All methods
have biases and limitations and we must use complementary methods to
obtain robust results.

@@

\textbf{\emph{Human samples.}} Fecal samples were obtained from 4
individuals who were part of an earlier study (44). These samples were
collected using a protocol approved by the University of Michigan
Institutional Review Board. For this study, the samples were
de-identified. DNA was extracted from the fecal samples using the
MOBIO\textsuperscript{TM} PowerMag Microbiome RNA/DNA extraction kit
(now Qiagen, MD, USA).

@@

\textbf{\emph{Sequence processing.}} The mothur software program (v
1.41) was used for all sequence processing steps (45). The protocol has
been previously published (18)
(\url{https://www.mothur.org/wiki/MiSeq_SOP}). Briefly, paired reads
were assembled using mothur's make.contigs command to correct errors
introduced by sequencing (18). Any assembled contigs that contained an
ambiguous base call, mapped to the incorrect region of the 16S rRNA
gene, or appeared to be a contaminant were removed from subsequent
analyses. Sequences were further denoised using mothur's pre.cluster
command to merge the counts of sequences that were within 2 nt of a more
abundant sequence. The VSEARCH implementation of UCHIME was used to
screen for chimeras (35, 46). At various stages in the sequence
processing pipeline for the mock community data, the mothur seq.error
command was used to quantify the sequencing error rate as well as the
true chimera rate. This command uses the true sequences from the mock
community to generate all possible chimeras and removes any contigs that
were at least three bases more similar to a chimera than to a reference
sequence. The command then counts the number of substitutions,
insertions, and deletions in the contig relative to the reference
sequence and reports the error rate without the inclusion of chimeric
sequences (19). The reference sequences and operon copy number for each
bacterium were obtained from the ZymoBIOMICS\textsuperscript{TM}
Microbial Community DNA Standard protocol
(\url{https://web.archive.org/web/20181221151905/https://www.zymoresearch.com/media/amasty/amfile/attach/_D6305_D6306_ZymoBIOMICS_Microbial_Community_DNA_Standard_v1.1.3.pdf}).
Sequences were assigned to operational taxonomic units (OTUs) at a
threshold of 3\% dissimilarity using the OptiClust algorithm (47). To
adjust for unequal sequencing when measuring alpha and beta diversity,
all samples were rarefied to 1,000 sequences 1,000 times for downstream
analysis. The Good's coverage for the samples at this level of sampling
was routinely greater than 95\%.

\textbf{\emph{Statistical analysis.}} All analysis was done with the R
(v 3.5.0) software package (48). Data transformation and graphing were
completed using the tidyverse package (v 1.2.1) (49). The distance
matrix data was analyzed using the adonis function within the vegan
package (v 2.5.3) (50).

@@
